### 2022-07-20 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@2938f44d869e03e1f18265e29ad281d36f52bd1b 🚀

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. Food-web properties that can be derived
using MaxEnt are varied and pertain to different elements of the network
(i.e.~at the species (node), the interaction (edge) or the community
(network) levels). For example, at the species level, MaxEnt can be used
to predict the distribution of trophic levels among species, as well as
the distribution of species' vulnerability (number of predators) and
generality (number of preys). By contrast, at the interaction level,
predictions can be made on the distribution of interaction strengths in
weighted food webs. At the community level, it can give us a probability
distribution of potential networks or of specific measures of their
emerging structure. The applicability and performance of this method
depend on the ecological information available and on our capacity to
find the set of state variables that best represent natural systems and
to translate them into appropriate statistical constraints. MaxEnt has
been used in a wide range of disciplines, from thermodynamics to
chemistry and biology (Martyushev and Seleznev 2006). It has also been
proven useful in ecology, e.g.~in species distribution models (Phillips,
Anderson, and Schapire 2006) and macroecological models (Harte et al.
2008; Harte and Newman 2014). Maximizing a measure of entropy ensures
that the derived probability distributions are unique and least biased
under the set of constraints used. These constraints are built using
state variables, i.e.~variables that represent the macrostate of the
system. Having a validated maximum entropy model for the system at hand
allows us to make rigorous predictions using a minimal amount of data,
as well as helping us describe the most important factors driving that
system.

Despite its extensive use in graph and network theories (e.g., Park and
Newman 2004; van der Hoorn, Lippner, and Krioukov 2018), MaxEnt has in
comparison been little used in network ecology. Like other real system
networks, ecological networks are represented mathematically as graphs.
However, the very nature of ecological networks (directed simple graphs
frequently having self-loops) makes the mathematical optimization of
maximum entropy graph models more complicated than with many other types
of (non-ecological) networks. MaxEnt has nevertheless been used to
predict the degree distribution of bipartite ecological networks from
the number of species and the number of interactions (Williams 2011) and
to predict interaction strengths between species pairs using their
relative abundances within an optimal transportation theory regularized
with information entropy (Stock, Poisot, and De Baets 2021). However, to
the best of our knowledge, MaxEnt has never been used to predict
food-web structure directly, even though food webs are among the most
documented and widespread ecological networks.

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution (a
probability distribution) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second (a species' degree is the sum of
its in and out-degrees). Because of the scarcity of empirical data on
the number of interactions in ecological networks, we present a method
to predict \(L\) from \(S\) (Box 1), thus allowing the prediction of the
joint degree distribution from \(S\) solely. In turn, the second
approach consists in finding, using different constraints, the adjacency
matrix of maximum entropy from which network properties can be measured.
To do so, we used a flexible and heuristic approach based on simulated
annealing to find networks \emph{close} to maximum entropy. As discussed
above, our choice of algorithm stands from the very nature of food webs
(i.e.~simple directed networks allowing self-loops) that makes the
analytical derivation of a maximum entropy graph model difficult. We
first built our type I MaxEnt network model constrained by the
connectance of the network (i.e.~the ratio \(L/S^2\)). A comparison of
this model against empirical data indicated that connectance alone was
not sufficient to predict many aspects of network structure. For this
reason, we built our type II MaxEnt network model, which instead uses
the whole joint degree sequence as a constraint. Overall, we found that
this second model was much better at predicting food-web structure than
the one based on connectance.

@@

Entropy measures the amount of information given by the outcome of a
random variable. Many measures of entropy have been developed in physics
(Beck 2009), but only a fraction of them could be used as an
optimization measure with the principle of maximum entropy. According to
Beck (2009) and Khinchin (2013), a measure of entropy \(H\) should
satisfy four properties in the discrete case: (1) it should be a
function of a probability distribution \(p(n)\) only; (2) it should be
maximized when \(p(n)\) is uniform; (3) it should not be influenced by
outcomes with a null probability; and (4) it should be independent of
the order of information acquisition. The Shannon's entropy (Shannon
1948)

@@

where \(g_i\) is the mathematical formulation of the constraint \(i\)
and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
maximized by setting to \(0\) its partial derivative with respect to
\(p(n)\). We will show how this can be done when we derive the joint
degree distribution analytically from the number of species and the
number of interactions in food webs.

@@

We tested our MaxEnt models (both approaches) against open food-web data
queried from three different sources and integrated into what we call
our \emph{complete dataset}. First, all food webs archived on
\texttt{mangal.io} (Poisot et al. 2016; Banville, Vissault, and Poisot
2021) were directly queried from the database (\(n = 235\)). Most
ecological networks archived on Mangal are multilayer networks,
i.e.~networks that describe different types of interactions. We kept all
networks whose interactions were mainly of predation and herbivory
types, and removed the largest network (\(S = 714\)) for computational
efficiency reasons. Then, to this set we added food webs from two
different sources: the New-Zealand dataset (\(n = 21\); Pomeranz et al.
2018) and the Tuesday lake dataset (\(n = 2\); Cohen, Jonsson, and
Carpenter 2003). Of these two datasets, \(19\) networks had data on
species' relative abundances that were used in the neutral model
presented in a following subsection. These networks are part of what we
call our \emph{abundance dataset}, which is a subset of our complete
dataset.

@@

Our maximum entropy network models (second approach only) were compared
with two topological null models. The first is the type I null model of
Fortuna and Bascompte (2006), in which the probability that a species
\(i\) predates on another species \(j\) is given by

@@

\hypertarget{heuristical-models-networks-of-maximum-entropy}{%
\section{Heuristical models: Networks of maximum
entropy}\label{heuristical-models-networks-of-maximum-entropy}}

@@

We built two types of MaxEnt network models: one based on connectance
(type I MaxEnt network model) and the other based on the joint degree
sequence (type II MaxEnt network model). They are based on the same
constraints as the types I and II null models presented above. For both
models, we used a simulated annealing algorithm with \(4\) chains,
\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
first generated one random Boolean matrix with the same order (number of
species) as empirical webs, while maintaining the total number of
interactions (type I MaxEnt network model) or rows and columns sums
(type II MaxEnt network model). These are our initial configurations.
Then, we swapped interactions sequentially while maintaining the
original connectance or the joint degree sequence for types I and II
MaxEnt network model, respectively. Configurations with a higher SVD
entropy than the previous one in the chain were always accepted, whereas
they were accepted with a probability conditional to a decreasing
temperature when lower. The final configuration with the highest SVD
entropy among the four chains constitute our estimated MaxEnt network.
Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains after burn-in. For each network in our complete
and abundance datasets, we estimated their configuration with maximum
entropy using both types of MaxEnt network models.

>>>>

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. Food-web properties that can be derived
using MaxEnt are varied and pertain to different elements of the network
(i.e.~at the species (node), the interaction (edge) or the community
(network) levels). Because MaxEnt is a method of finding probability
distributions that are least biased given partial knowledge about a
system, these properties need to be represented probabilistically.
Otherwise, some data transformation might be needed. For example, at the
species level, MaxEnt can be used to predict the distribution of trophic
levels among species, as well as the distribution of species'
vulnerability (number of predators) and generality (number of preys). By
contrast, at the interaction level, predictions can be made on the
distribution of interaction strengths in weighted food webs. At the
community level, it can give us a probability distribution of potential
networks or of specific measures of their emerging structure. The
applicability and performance of this method depend on the ecological
information available and on our capacity to find the set of state
variables that best represent natural systems and to translate them into
appropriate statistical constraints. MaxEnt has been used in a wide
range of disciplines, from thermodynamics to chemistry and biology
(Martyushev and Seleznev 2006). It has also been proven useful in
ecology, e.g.~in species distribution models (Phillips, Anderson, and
Schapire 2006) and macroecological models (Harte et al. 2008; Harte and
Newman 2014). The probability distribution of the variable of interest
with the maximum entropy value is unique and least biased under the set
of constraints used. These constraints are built using state variables,
i.e.~variables that represent the macrostate of the system. Having a
validated maximum entropy model for the system at hand allows us to make
rigorous predictions using a minimal amount of data, as well as helping
us describe the most important factors driving that system.

Despite its extensive use in graph and network theories (e.g., Park and
Newman 2004; van der Hoorn, Lippner, and Krioukov 2018), MaxEnt has in
comparison been little used in network ecology. Like other real system
networks, food webs and other ecological networks are represented
mathematically as graphs. However, the very nature of food webs
(directed simple graphs frequently having self-loops) makes the
mathematical optimization of maximum entropy graph models more
complicated than with many other types of (non-ecological) networks.
MaxEnt has nevertheless been used to predict the degree distribution of
bipartite ecological networks from the number of species and the number
of interactions (Williams 2011) and to predict interaction strengths
between species pairs using their relative abundances within an optimal
transportation theory regularized with information entropy (Stock,
Poisot, and De Baets 2021). However, to the best of our knowledge,
MaxEnt has never been used to predict food-web structure directly, even
though food webs are among the most documented and widespread ecological
networks.

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution
(the joint probability distribution that a species has a given number of
prey and predators in a food web) of maximum entropy using only the
number of species \(S\) and the number of interactions \(L\) as state
variables. Then, we predicted the degree distribution of maximum entropy
directly from the joint degree distribution since the first is the sum
of the marginal distributions of the second (a species' degree is the
sum of its in and out-degrees). Because of the scarcity of empirical
data on the number of interactions in ecological networks, we present a
method to predict \(L\) from \(S\) (Box 1), thus allowing the prediction
of the joint degree distribution from \(S\) solely. In turn, the second
approach consists in finding, using different constraints, the adjacency
matrix of maximum entropy from which network properties can be measured.
To do so, we used a flexible and heuristic approach based on simulated
annealing (an optimization algorithm) to find networks \emph{close} to
maximum entropy. As discussed above, our choice of algorithm stands from
the very nature of food webs (i.e.~simple directed networks allowing
self-loops) that makes the analytical derivation of a maximum entropy
graph model difficult. We first built our type I MaxEnt network model
constrained by the connectance of the network (i.e.~the ratio
\(L/S^2\)). A comparison of this model against empirical data indicated
that connectance alone was not sufficient to predict many aspects of
network structure. For this reason, we built our type II MaxEnt network
model, which instead uses the whole joint degree sequence as a
constraint. Overall, we found that this second model was much better at
predicting food-web structure than the one based on connectance.

@@

The least-biased probability distribution given the constraints used is
the one with the highest entropy among all probability distributions
that satisfy these constraints. Entropy is a measure of the average
amount of information given by the outcome of a random variable. Many
measures of entropy have been developed in physics (Beck 2009), but only
a fraction of them could be used as an optimization measure with the
principle of maximum entropy. According to Beck (2009) and Khinchin
(2013), a measure of entropy \(H\) should satisfy four properties in the
discrete case: (1) it should be a function of a probability distribution
\(p(n)\) only; (2) it should be maximized when \(p(n)\) is uniform; (3)
it should not be influenced by outcomes with a null probability; and (4)
it should be independent of the order of information acquisition. The
Shannon's entropy (Shannon 1948)

@@

where \(g_i\) is the mathematical formulation of the constraint \(i\)
and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
maximized by setting to \(0\) its partial derivative with respect to
\(p(n)\). In this contribution, we show how this can be done by deriving
the joint degree distribution analytically from the number of species
and the number of interactions in food webs.

@@

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were directly
queried from the database (\(n = 235\)). Most ecological networks
archived on Mangal are multilayer networks, i.e.~networks that describe
different types of interactions. We kept all networks whose interactions
were mainly of predation and herbivory types, and removed the largest
network (\(S = 714\)) for computational efficiency reasons. Then, to
this set we added food webs from two different sources: the New-Zealand
dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday lake dataset
(\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these two datasets,
\(19\) networks had data on species' relative abundances that were used
in the neutral model (Box 2). These \(19\) networks constitute our
\emph{abundance dataset}, which is a subset of our complete dataset.

@@

Our heuristic MaxEnt models, predicting adjacency matrices, were
compared with two topological null models. The first is the type I null
model of Fortuna and Bascompte (2006), in which the probability that a
species \(i\) predates on another species \(j\) is given by

@@

\hypertarget{heuristic-models-networks-of-maximum-entropy}{%
\section{Heuristic models: Networks of maximum
entropy}\label{heuristic-models-networks-of-maximum-entropy}}

@@

We built two types of MaxEnt network models: one based on connectance
(type I MaxEnt network model) and the other based on the joint degree
sequence (type II MaxEnt network model). They are based on the same
constraints as the types I and II null models presented above. For both
models, we used a simulated annealing algorithm with \(4\) chains,
\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
first generated one random Boolean matrix with the same order (number of
species) as empirical webs, while maintaining the total number of
interactions (type I MaxEnt network model) or rows and columns sums
(type II MaxEnt network model). These are our initial configurations.
Then, we swapped interactions sequentially while maintaining the
original connectance or the joint degree sequence for types I and II
MaxEnt network model, respectively. Configurations with a higher SVD
entropy than the previous one in the chain were always accepted, whereas
they were accepted with a probability conditional to a decreasing
temperature when lower. The final configuration with the highest SVD
entropy among the four chains constitute our estimated MaxEnt network.
Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains (or part of the chains). For each network in our
complete and abundance datasets, we estimated their configuration with
maximum entropy using both types of MaxEnt network models.
